### 2021-09-24 revision/ch5langevin.tex
getting close.

Bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Owing to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (secs. \ref{sec:birthdeath}-\ref{sec:renewal}).
Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (sec. \ref{sec:birthdeath}).
This formulation requires the instantaneous velocity distributions of sediment particles \citep[e.g.][]{Ancey2020a}, yet unfortunately, understanding of these distributions remains limited.
There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in sec. \ref{sec:langevin} of the introduction describe some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe all experimental observations reported to date \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.

@@

Stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec. \ref{sec:langevin}). \citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces into a simplified velocity-dependent force, again modeling fluctuations with Gaussian white noise.
Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
A physical description for the full range of experimentally-observed bedload velocity distributions remains an elusive target.

In this chapter I explore the possibility that the particle velocity distribution is controlled by momentum dissipation from particle-bed collisions by incorporating ideas from the physics of granular media.
A major paradigm in granular dynamics is to describe particles as a continuum or "granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in ch. \ref{ch:Introduction} such a continuum assumption is unlikely to be satisfied for weak bedload transport conditions. Instead, particles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}. 
In the theory of rarefied granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970,Briliantov2004}.
Collisions in these models involve episodic kicks to the particle momenta at random times, usually parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
Such episodic forcing is distinct from the smooth friction terms included in the current stochastic models describing bedload particle velocities.

Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport which replaces the smooth friction terms of earlier models by episodic particle-bed collisions.
The driving motivation is to test the hypothesis that particle-bed collision characteristics can explain the range of experimentally-observed streamwise bed load particle velocity distributions.
A secondary motivation is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
I develop the model and explain its key assumptions in sec. \ref{sec:langmodel}. Then I present the analytical solution of the model and other major results in \ref{sec:langresults}. Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in secs. \ref{sec:langdiscussion} and \ref{sec:langconclusion}.

@@

Figure \ref{fig:fig1} indicates the configuration of the model in this chapter.
Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady flow.
The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Jackson1982, Warburton1992,Hassan2001}. Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles, but rarely with other moving particles \citep[cf.][]{Williams2020}.

Particles respond to turbulent flow forces and episodic particle-bed collision forces. In contrast to the computational physics approach, I do not aim to characterize the exact time series of the forces on an individual particle. Instead, I model the ensemble of possible force time series that particles could conceivably experience. Each force time series realization implies a different velocity time series $u(t)$ in the downstream direction.
The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
Here, the description of bedload particle velocity fluctuations will be more detailed than ch. \ref{ch:flux}, but entrainment and deposition processes will be neglected, meaning the model of this chapter is only applicable to moving particles.
I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.

\subsection{Episodic collision forces}
Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid in the interstices between approaching particles \citep{Joseph2001,Schmeeckle2001}. 
The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of resitution may be less relevant than other factors such as collision geometry and bed deformation \citep{}. Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in fig. \ref{fig:fig1}. This elasticity characterizes the fraction of downstream momentum lost in a collision. If the streamwise velocity just prior to a collision is $u$, just after the collision it becomes $\varepsilon u$. The elasticity ranges from $\varepsilon=0$, representing a completely inelastic collision, to $\varepsilon=1$, representing a completely elastic collision.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig1Concept.png}}
	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces. During saltation, pre-collisional streamwise velocities $u$ are transformed to postcollisional velocities $\ve u < u$.}
	\label{fig:fig1}
\end{figure}

@@

Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (sec. \ref{sec:einflux}):
\be F_C(u,t) = - m u \sum_{k=1}^{N_\nu(t)}(1-\varepsilon_k)\delta(t-\tau_k). \label{eq:col} \ee
Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
The time intervals $\Delta \tau_k = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau).$
This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momenta. This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions. These conditions should be satisfied for idealized saltation trajectories as depicted in fig. \ref{fig:fig1}.

@@

Our understanding of the fluid forces remains incomplete.
Existing formulas were derived in the absence of boundaries, so it is unclear how these formulas should be modified near the bed.
Even ignoring this complication, the added mass and Basset forces are difficult to compute and therefore impractical to include in analytical models.
These challenges require researchers to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michalades1997,Armenio2001}.
As a result, the majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the primary force(s) driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.

The downstream drag force with the $\Rey_p$ correction is usually written 
\be F_D = \frac{\pi}{8} \rho_f d^2 C_D(\Rey_p) |V|V, \label{eq:draggy}\ee
where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
Experiments on stationary particles indicate that drag forces fluctuate rapidly and lie on wide Gaussian-like distributions \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities should be expected to vary due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
The nonlinearity in eq. \ref{eq:draggy} will amplify these slip velocity variations, generating large and erratic drag force fluctuations.

@@

With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics in the downstream direction becomes
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
Mathematically, eq. \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce velocity jumps while turbulence generates velocity diffusion. The collision term is multiplicative in the sense that $u$ multiplies the Poisson noise $\xi_{\nu,\ve}(t)$.

Models like eq. \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning eq. \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by eq. \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

@@

The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
In this equation, I introduced the parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$. The term
\be I_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.

Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Nonlocality is introduced by the collision integral (eq. \ref{eq:colint}) which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random restitution coefficients \citep[cf.][]{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $(p = m u)$, considering them as uncorrelated random variables \citep[cf.][]{Feller1968}.

Owing to its nonlocality, eq. \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
For bedload particle collisions with the bed, we can expect a mode elasticity since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.

@@

\subsection{Derivation of the bedload velocity distribution}
\label{sec:langsolution}
Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
With the truncation to two terms, eq. \ref{eq:master} gives 
\be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
which is an advanced functional differential equation. This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
Such equations have seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.

In the appendix sec. \ref{sec:langsteadyderiv} eq. \ref{eq:governer} is solved with Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(- \tilde{D} \lambda_-^2 \ve^{-2m} + \tilde{\Gamma} \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in appendix eq. \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors $K_\pm$ are 
\be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values. The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.

It is possible to derive the moments of this probability distribution by multiplying eq. \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[cf.][]{Cox1965}.
These calculations are provided in appendix sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}. \label{eq:meanu}\ee
This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
The second moment is
\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
\be \sigma_u^2 = \frac{2 \tilde{D}+ \tilde{\Gamma}^2}{1-\ve^2}. \label{eq:varu}\ee
This result demonstrates that within the model, velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
This is consistent with the obvious similarities between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
In contrast, particle velocity fluctuations in eq. \ref{eq:varu} depend sharply on the parameters representing particle-bed collisions, suggesting collisions could be the leading control on particle velocity fluctuations.

>>>>

Bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Owing to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (Secs. \ref{sec:birthdeath}-\ref{sec:renewal}).
Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (Sec. \ref{sec:birthdeath}).
This formulation requires the instantaneous velocity distributions of sediment particles \citep[e.g.][]{Ancey2020a}, yet unfortunately, understanding of these distributions remains limited.
There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in Sec. \ref{sec:langevin} of the introduction describe some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe all experimental observations reported to date \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.

@@

Stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (Sec. \ref{sec:langevin}). \citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces into a simplified velocity-dependent force, again modeling fluctuations with Gaussian white noise.
Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
A physical description for the full range of experimentally-observed bedload velocity distributions remains an elusive target.

In this chapter I explore the possibility that the particle velocity distribution is controlled by momentum dissipation from particle-bed collisions by incorporating ideas from the physics of granular media.
A major paradigm in granular dynamics is to describe particles as a continuum or "granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in Ch. \ref{ch:Introduction} such a continuum assumption is unlikely to be satisfied for weak bedload transport conditions. Instead, particles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}. 
In the theory of rarefied granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970,Brilliantov2004}.
Collisions in these models involve episodic kicks to the particle momenta at random times, usually parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
Such episodic forcing is distinct from the smooth friction terms included in the current stochastic models describing bedload particle velocities.

Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport which replaces the smooth friction terms of earlier models by episodic particle-bed collisions.
The driving motivation is to test the hypothesis that particle-bed collision characteristics can explain the range of experimentally-observed streamwise bed load particle velocity distributions.
A secondary motivation is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
I develop the model and explain its key assumptions in Sec. \ref{sec:langmodel}. Then I present the analytical solution of the model and other major results in \ref{sec:langresults}. Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in Secs. \ref{sec:langdiscussion} and \ref{sec:langconclusion}.

@@

Figure \ref{fig:fig1} indicates the configuration of the model in this chapter.
Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady flow.
The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Ashworth1989, Warburton1992}. Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles, but rarely with other moving particles \citep[cf.][]{Williams2021}.

Particles respond to turbulent flow forces and episodic particle-bed collision forces. In contrast to the computational physics approach, I do not aim to characterize the exact timeseries of the forces on an individual particle. Instead, I model the ensemble of possible force timeseries that particles could conceivably experience. Each force timeseries realization implies a different velocity timeseries $u(t)$ in the downstream direction.
The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
Here, the description of bedload particle velocity fluctuations will be more detailed than Ch. \ref{ch:flux}, but entrainment and deposition processes will be neglected, meaning the model of this chapter is only applicable to moving particles.
I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.

\subsection{Episodic collision forces}
Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid between approaching particles \citep{Joseph2001,Schmeeckle2001}. 
The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of resitution may be less relevant than other factors such as collision geometry and bed deformation. Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in Fig. \ref{fig:fig1}. This elasticity characterizes the fraction of downstream momentum lost in a collision. If the streamwise velocity just prior to a collision is $u$, just after the collision it becomes $\varepsilon u$. The elasticity ranges from $\varepsilon=0$, representing a completely inelastic collision, to $\varepsilon=1$, representing a completely elastic collision.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig1Concept.png}}
	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces. During saltation, pre-collisional streamwise velocities $u$ are transformed to post-collisional velocities $\ve u < u$.}
	\label{fig:fig1}
\end{figure}

@@

Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (Sec. \ref{sec:einflux}):
\be F_C(u,t) = - m u \sum_{k=1}^{N_\nu(t)}(1-\varepsilon_k)\delta(t-\tau_k). \label{eq:col} \ee
Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
The time intervals $\Delta \tau_k = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau).$
This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momenta. This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions. These conditions should be satisfied for idealized saltation trajectories as depicted in Fig. \ref{fig:fig1}.

@@

Our understanding of the fluid forces remains incomplete.
Existing formulas were derived in the absence of boundaries, so it is unclear how these formulas should be modified near the bed.
Even ignoring this complication, the added mass and Basset forces are difficult to compute and therefore impractical to include in analytical models.
These challenges require researchers to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michaelides1997,Armenio2001}.
As a result, the majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the primary force(s) driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.

The downstream drag force with the $\Rey_p$ correction is usually written 
\be F_D = \frac{\pi}{8} \rho_f d^2 C_D(\Rey_p) |V|V, \label{eq:draggy}\ee
where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
Experiments on stationary particles indicate that drag forces fluctuate rapidly and lie on wide Gaussian-like distributions \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities should be expected to vary due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
The nonlinearity in Eq. \ref{eq:draggy} will amplify these slip velocity variations, generating large and erratic drag force fluctuations.

@@

With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics in the downstream direction becomes
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
Mathematically, Eq. \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce velocity jumps while turbulence generates velocity diffusion. The collision term is multiplicative in the sense that $u$ multiplies the Poisson noise $\xi_{\nu,\ve}(t)$.

Models like Eq. \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning Eq. \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by Eq. \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

@@

The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix Sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
In this equation, I introduced the parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$. The term
\be I_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.

Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Nonlocality is introduced by the collision integral (Eq. \ref{eq:colint}) which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random restitution coefficients \citep[cf.][]{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $(p = m u)$, considering them as uncorrelated random variables \citep[cf.][]{Feller1968}.

Owing to its nonlocality, Eq. \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
For bedload particle collisions with the bed, we can expect a mode elasticity since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.

@@

\subsection{Derivation of the bedload velocity distribution}
\label{sec:langsolution}
Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
With the truncation to two terms, Eq. \ref{eq:master} gives 
\be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
which is an advanced functional differential equation. This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
Such equations have seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.

In the appendix Sec. \ref{sec:langsteadyderiv} Eq. \ref{eq:governer} is solved with Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(- \tilde{D} \lambda_-^2 \ve^{-2m} + \tilde{\Gamma} \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in appendix Eq. \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors $K_\pm$ are 
\be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values. The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.

It is possible to derive the moments of this probability distribution by multiplying Eq. \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[cf.][]{Cox1965}.
These calculations are provided in appendix Sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}. \label{eq:meanu}\ee
This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
The second moment is
\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
\be \sigma_u^2 = \frac{2 \tilde{D}+ \tilde{\Gamma}^2}{1-\ve^2}. \label{eq:varu}\ee
This result demonstrates that within the model, velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
This is consistent with the obvious similarities between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
In contrast, particle velocity fluctuations in Eq. \ref{eq:varu} depend sharply on the parameters representing particle-bed collisions, suggesting collisions could be the leading control on particle velocity fluctuations.
